The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-15 paper/iphas-dr2.tex
First short at detection stats and Vega fluxes

E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, 
R. Greimel$^3$, \newauthor

@@

A. Mampaso$^{6,7}$, 
S. E. Sale$^8$, \newauthor

@@

J. J. Drake$^{10}$,
J. Eisl\"offel$^{11}$, \newauthor

@@

S. Greiss$^{13}$,
A. S. Hales$^{14}$, \newauthor

@@

@@

C. Knigge$^{15}$,
T. Kupfer$^{5}$,
D. J. Lennon$^{16}$,
J. R. Lewis$^{2}$,  \newauthor

@@

R. A. H. Morris$^{17}$,
T. Naylor$^{18}$,
Q. A. Parker$^{19,20,21}$, 
S. Phillipps$^{17}$, \newauthor

@@

L. Sabin$^{22}$,
S. Scaringi$^{23,24}$,
D. Steeghs$^{13}$, \newauthor
Y. C. Unruh$^{25}$, 

@@

K. Viironen$^{26}$, 
J. S. Vink$^{27}$, 
N. A. Walton$^{2}$, \newauthor

@@

A. A. Zijlstra$^{28}$.

@@

$^{15}$School of Physics \& Astronomy, University of Southampton, Southampton, SO17 1BJ, U.K.\\
(see the source file for affiliations 16 to 28)
%$^{16}$European Space Astronomy Centre (ESAC), Villafranca del Castillo, Villanueva de la Canada, E-28692 Madrid, Spain.\\
%$^{17}$School of Physics, Bristol University, Tyndall Avenue, Bristol, BS8 1TL, U.K.\\
%$^{18}$School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, U.K.\\
%$^{19}$Department of Physics and Astronomy, Macquarie University, Sydney, NSW 2109, Australia.\\
%$^{20}$Australian Astronomical Observatory, PO Box 915, North Ryde, NSW 1670, Australia.\\
%$^{21}$Research Centre for Astronomy, Astrophysics and Astrophotonics, Macquarie University, Sydney, NSW 2109 Australia.\\
%$^{22}$Instituto de Astonom\'ia y Meteorolog\'ia, Departamento de F\'isica, CUCEI, Universidad de Guadalajara, Mexico.\\
%$^{23}$Instituut voor Sterrenkunde, K.U. Leuven, Celestijnenlaan 200D, B-3001 Leuven, Belgium.\\
%$^{24}$Max-Planck-Institut f\"ur extraterrestrische Physik (MPE), Giessenbachstrasse 1, D-85748, Garching bei M\"unchen, Germany. \\
%$^{25}$Department of Physics, Blackett Laboratory, Imperial College London, Prince Consort Road, London, SW7 2AZ, U.K.\\
%$^{26}$Centro de Estudios de F\'isica del Cosmos de Arag\'on, Plaza San Juan 1, Planta 2, Teruel, 44001, Spain. \\
%$^{27}$Armagh Observatory, College Hill, Armagh, Northern Ireland, BT61 9DG, U.K.\\
%$^{28}$Jodrell Bank Centre for Astrophysics, School of Physics \& Astronomy, University of Manchester, Manchester M13 9PL, U.K.

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30$ to 215\degr\ 
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92 per cent of the footprint.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

Enhancement of the detection tables by creating new columns
is the necessary first step because the tables generated by the CASU pipeline
refer e.g. to source positions in pixel coordinates, to photometric measures 
in number counts etc., rather than in directly useful astronomical terms.
To transform these data into user-friendly quantities,
we have largely adopted the units and naming conventions
which are in use at the 
WFCAM Science Archive \citep[WSA;][]{Hambly2008}
and the VISTA Science Archive \citep[VSA;][]{Cross2012}.
These archives curate the high-resolution near-infrared photometry from both
the UKIDSS Galactic Plane Survey \citep[GPS;][]{Lawrence2007,Lucas2008}
and the 
VISTA Variables in the Via Lactea survey \cite[VVV;][]{Minniti2010}.
There is a significant degree of overlap
between the footprints of UKIDSS/GPS and IPHAS,
and hence by adopting a similar catalogue format
we hope to facilitate scientific applications
which combine both data sets.

@@

@@

The catalogue is made available through the Vizier
catalogue tool (http://vizier.u-strasbg.fr)
where it can be queried
(Conesearch/TAP URL TBD).
The catalogue can also be downloaded in its entirety
from the IPHAS website (www.iphas.org) as a collection 
of binary FITS tables occupying 50~GB,
together with a script
to ingest these tables into a PostgreSQL database.

The website also provides access to the pipeline-processed
imaging data, which we have updated to include
the re-calibrated zeropoints and the best-available
astrometric solution in the image headers.

@@

Instructions for acknowledging IPHAS can be found on our website.

@@

\section{Converting IPHAS magnitudes into fluxes}

\begin{table}[h]
    \caption{Wavelength and mean monochromatic flux density for Vega in the IPHAS filter system.}
    \label{tbl:flux}
    \begin{tabular}{lrrrr}
    \toprule
     Band & $\lambda_0$ & $\lambda_p$ & EW  & $f_\lambda$   \\
     & \AA & \AA & \AA & erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$  \\
    \midrule
$r$ &  6223  &   6211  &  785.6  &  2.47e-9  \\
\ha &  6568  &   6568  &   59.6  &  1.81e-9  \\
$i$ &  7674  &   7674  &  759.9  &  1.30e-9   \\
    \bottomrule
    \end{tabular}
\end{table}

>>>>

E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, \newauthor
R. Greimel$^3$, 

@@

A. Mampaso$^{6,7}$, \newauthor
S. E. Sale$^8$, 

@@

J. J. Drake$^{10}$, \newauthor
J. Eisl\"offel$^{11}$, 

@@

S. Greiss$^{13}$, \newauthor
A. S. Hales$^{14}$, 

@@

L. Huckvale$^{15}$,

@@

C. Knigge$^{16}$,
T. Kupfer$^{5}$, \newauthor
D. J. Lennon$^{17}$,
J. R. Lewis$^{2}$,  

@@

R. A. H. Morris$^{18}$,
T. Naylor$^{19}$, \newauthor
Q. A. Parker$^{20,21,22}$, 
S. Phillipps$^{17}$, 

@@

L. Sabin$^{23}$, \newauthor
S. Scaringi$^{24,25}$,
D. Steeghs$^{13}$, 
Y. C. Unruh$^{26}$, 

@@

K. Viironen$^{27}$, \newauthor
J. S. Vink$^{28}$, 
N. A. Walton$^{2}$, 

@@

A. A. Zijlstra$^{15}$.

@@

$^{15}$Jodrell Bank Centre for Astrophysics, School of Physics \& Astronomy, University of Manchester, Manchester M13 9PL, U.K.\\
%$^{16}$School of Physics \& Astronomy, University of Southampton, Southampton, SO17 1BJ, U.K.\\
%$^{17}$European Space Astronomy Centre (ESAC), Villafranca del Castillo, Villanueva de la Canada, E-28692 Madrid, Spain.\\
%$^{18}$School of Physics, Bristol University, Tyndall Avenue, Bristol, BS8 1TL, U.K.\\
%$^{19}$School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, U.K.\\
%$^{20}$Department of Physics and Astronomy, Macquarie University, Sydney, NSW 2109, Australia.\\
%$^{21}$Australian Astronomical Observatory, PO Box 915, North Ryde, NSW 1670, Australia.\\
%$^{22}$Research Centre for Astronomy, Astrophysics and Astrophotonics, Macquarie University, Sydney, NSW 2109 Australia.\\
%$^{23}$Instituto de Astonom\'ia y Meteorolog\'ia, Departamento de F\'isica, CUCEI, Universidad de Guadalajara, Mexico.\\
%$^{24}$Instituut voor Sterrenkunde, K.U. Leuven, Celestijnenlaan 200D, B-3001 Leuven, Belgium.\\
%$^{25}$Max-Planck-Institut f\"ur extraterrestrische Physik (MPE), Giessenbachstrasse 1, D-85748, Garching bei M\"unchen, Germany. \\
%$^{26}$Department of Physics, Blackett Laboratory, Imperial College London, Prince Consort Road, London, SW7 2AZ, U.K.\\
%$^{27}$Centro de Estudios de F\'isica del Cosmos de Arag\'on, Plaza San Juan 1, Planta 2, Teruel, 44001, Spain. \\
%$^{28}$Armagh Observatory, College Hill, Armagh, Northern Ireland, BT61 9DG, U.K.\\
$^{16-28}$ (affiliations 16 to 28 are given in the source file)

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30\degr$ to 215\degr\ 
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92 per cent of the footprint.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

Enhancement of the detection tables by creating new columns
is the necessary first step because the tables generated by the CASU pipeline
refer e.g. to source positions in pixel coordinates, to photometric measures 
in number counts etc., rather than in directly useful astronomical terms.
To transform these data into user-friendly quantities,
we have largely adopted the units and naming conventions
which are in use at the 
WFCAM Science Archive \citep[WSA;][]{Hambly2008}
and the VISTA Science Archive \citep[VSA;][]{Cross2012}.
These archives curate the high-resolution near-infrared photometry from both
the UKIRT Infrared Deep Sky Survey \citep[UKIDSS;][]{Lawrence2007}
and the 
VISTA Variables in the Via Lactea survey \cite[VVV;][]{Minniti2010}.
There is a significant degree of overlap
between the footprints of UKIDSS Galactic Plane Survey \citep[GPS;][]{Lucas2008}
and IPHAS,
and hence by adopting a similar catalogue format
we hope to facilitate scientific applications
which combine both data sets.

@@

\begin{table}
    \caption{Detection statistics.}
    \label{tbl:detections}
        \begin{tabular}{lrcrrr}
        \toprule
        Detection & N & \%($nObs>1$) &
        $\overline{pStar}$ & $\overline{(r-i)}$ & $\overline{(r-\ha)}$ \\
        \midrule
        $r$, $i$, \ha & 159 (73\%) & 91\% & 0.82 & 1.03 & 0.44  \\
        $r$, $i$ (not \ha) & 25 (11\%) & 77\% & 0.70 &  1.26 & - \\
        $i$, \ha\ (not $r$) & 3 (1\%) & 73\% & 0.49 &  & \\
        $r$, \ha\ (not $i$) & 2 (1\%) & 65\% & 0.53 & - & 0.58 \\
        only $i$ & 15 (7\%) & 43\% & 0.44 \\
        only $r$ & 9 (4\%) & 27\% & 0.42 \\
        only \ha & 6 (3\%) & 12\% & 0.11 \\
        \midrule
        Total & 219 & 81\% \\
        \bottomrule
        \end{tabular}
\end{table}

% riha: 158831809 / 144339044
% ri: 24558323 / 18878201
% iha: 3422905 / 2495655
% rha: 1606867 / 1036607
% i: 15463660 / 6612351
% r: 9263422 / 2478774
% ha: 5844538 / 727732

@@

The catalogue is made available on our website (www.iphas.org)
as a collection of binary FITS tables,
each covering a $5\degr\times5\degr$ tile
of the survey footprint.
The catalogue can be queried in two ways.
First, it is made available through the
Vizier on-line service (http://vizier.u-strasbg.fr)
where it can be queried using their web interface
or using Virtual Observatory (VO) protocols.
Second, on our website we provide a script
to ingest the binary FITS tables into a
{\sc PostgreSQL} database on a user's computer.

The website also provides access to the pipeline-processed
imaging data, which we have updated to include
a new keyword {\sc photzp},
which contains the re-calibrated zeropoint.
It can be used to convert the number counts $DN$
-- which are the pixel values -- 
into Vega-based magnitudes $m$ using:
\begin{equation}
\begin{split}
   m  = \textsc{photzp} - 2.5 \log_{10}(DN)
\label{eqn:mag}
\end{split}
\end{equation}
That is, the {\sc photzp} has been computed such that it
absorbs the required corrections for atmospheric extinction,
gain variations, exposure time, and the re-calibration shift
derived in this work.

To convert the magnitude $m$ into fluxes,
we list in Table~\ref{tbl:flux} something.

% and the best-available
%astrometric solution in the image headers.

@@

Users of the data are reminded to cite both \citet{Drew2005}
and this paper.

\begin{table}
    \caption{Wavelength and mean monochromatic flux density for Vega in the IPHAS filter system.}
    \label{tbl:flux}
    \begin{tabular}{lrrrr}
    \toprule
     Band & $\lambda_0$ & $\lambda_p$ & EW  & $f_\lambda$   \\
     & \AA & \AA & \AA & erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$  \\
    \midrule
$r$ &  6223  &   6211  &  785.6  &  2.47e-9  \\
\ha &  6568  &   6568  &   59.6  &  1.81e-9  \\
$i$ &  7674  &   7674  &  759.9  &  1.30e-9   \\
    \bottomrule
    \end{tabular}
\end{table}

@@
